### 2016-01-13 docs/paper.tex
Paper: Init commit

% vim:set spell tw=79:

\documentclass[article]{uibk}
\title{Exploitation Techniques and Mitigations}
\author{Alex Hirsch \and Patrick Ober}
\date{2016-01-31}

\begin{document}

\maketitle

% \section*{Abstract}
% \label{sec:abstract}

% \tableofcontents

% \newpage

\section*{Acknowledgement}

A university course at Rensselaer Polytechnic
Institut\footnote{\url{http://rpi.edu/}} held in Spring 2015 focused on
\textit{Modern Binary Exploitation}. They made their course material available
on GitHub \cite{RPISEC} under the Creative Commons Attribution-NonCommercial
4.0 International
license\footnote{\url{https://creativecommons.org/licenses/by-nc/4.0/legalcode}}.
We reused a lot of their material in this project.

We highly recommend checking them out and having a look at their material for
further details.

\section{Introduction}

Exploiting binaries was comparatively easy in the early days of computing.
Usually there were no special mitigation techniques in place trying to prevent
even the most simplest exploits. This is the point in time where we will start
of. First we talk about two very simple exploits, namely the Format String
Exploit and the Buffer Overflow in combination with Shell Code. Note that there
is a huge collection of exploitation techniques known to the public and we will
thereby only look at a very small fraction of them.

But before we can introduce these two exploits, some background knowledge is
required. This will be handled by the next section, which provides a short
overview of the relevant components in our target architecture, the x86
platform.

After that both techniques are introduced to the reader, followed by the first
mitigation technique, Data Execution Prevention (DEP). From there on we will
keep on using the buffer overflow technique with some adaptations to circumvent
DEP. At this point Return Oriented Programming (ROP) is introduced.

This directly leads to Address Space Layout Randomization (ASLR) the next
mitigation mechanism we will discuss. Again the buffer overflow technique can
be adapted to break ASLR through the use of additional information.

Since neither DEP nor ASLR provide significant protection against even this
simple technique, an additional mitigation is put into place in the form of
Stack Cookies.

Examples will be provided along the way to support the reader and provide some
additional explanation.

Control Flow Integrity (CFI), Heap Corruption and polymorphic code will follow
in a more compressed manner to communicate the main idea behind each of them.

Finally we will conclude with a word about other architectures (x86\_64 and
ARM) and a lookout that even languages considered secure have their own set of
exploitation techniques an attacker could leverage.

\section{Platform x86}

This section will teach necessary background knowledge about the target
platform to fully conceive the following techniques. But first let us elaborate
why x86 has been chosen in the first place.

At the time these techniques (and the related mitigations) were established,
x86 was the most common platform. Since most exploits easily translate over
from x86 to other architectures, especially x86\_64 which very common nowadays.
Also, most material found on the internet regarding this and related topics
cover x86.

More detailed explanations can be found on
Wikipedia\footnote{\url{https://en.wikipedia.org/wiki/X86}} or the Intel
Manual\footnote{\url{https://www-ssl.intel.com/content/www/us/en/processors/architectures-software-developer-manuals.html}}.

\subsection{CPU and registers}

\subsection{Memory}

\newpage

\bibliography{references}

\end{document}

### 2016-01-14 docs/paper.tex
Paper: Add section about main assumption

@@
 
+\subsection{Main Assumption}
+
+Throughout this work we assume that we know the target binary (and the
+libraries it uses). Let us show that this assumption is quite reasonable to
+make by looking through the eyes of the adversary. An attacker who wants to
+penetrate a target machine and get control over it would most likely choose the
+easiest path, by exploiting the weakest link. Most machines relevant to an
+attackers interest will run provide multiple services. For example, while the
+main server of a small business company may run a homemade communication server
+for interaction between them and their clients, it may also run a standard web
+server. Sending a misspelled request to the server may lead following response:
+
+\begin{pre*}{autogobble,xleftmargin=1cm}
+    <!DOCTYPE HTML PUBLIC "-//IETF//DTD HTML 2.0//EN">
+    <html><head>
+    <title>400 Bad Request</title>
+    </head><body>
+    <h1>Bad Request</h1>
+    <p>Your browser sent a request that this server could not understand.<br />
+    </p>
+    <hr>
+    <address>Apache/2.2.22 (Ubuntu) Server at ovinnik.canonical.com Port 80</address>
+    </body></html>
+    Connection closed by foreign host.
+\end{pre*}
+
+The web server tells us his exact version and since it also provides
+information about the operating system an attacker can easily copy the basic
+setup to test and tweak his exploits.
+
 \section{Platform x86}

### 2016-01-14 overflow/overflow.c
files used at presentation

@@ -1,0 +1,19 @@
+#include <stdio.h>
+
+void mordor(){
+	printf("One does not simply jump into mordor()!\n");
+}
+
+void echo(){
+	char buffer[20];
+	printf("Enter text:\n");
+	scanf("%s", buffer);
+	printf("You entered: %s\n", buffer);
+}
+
+int main(){
+
+	echo();
+	
+	return 0;
+}

### 2016-01-14 rop/rop.c

@@ -1,0 +1,19 @@
+#include <stdio.h>
+#include <string.h>
+
+char* not_used = "/bin/sh";
+
+void not_called(){
+		printf("Not quite a shell...\n");
+		system("/bin/date");
+}
+
+void vulnerable_function(char* string){
+	char buffer[100];
+	strcpy(buffer, string);
+}
+
+int main(int argc, char** argv){
+		vulnerable_function(argv[1]);
+		return 0;
+}

### 2016-01-14 shellinjection/vuln.c

@@ -1,0 +1,13 @@
+#include <stdio.h>
+#include <string.h>
+
+void func(char *name){
+	char buf[100];
+	strcpy(buf, name);
+	printf("Welcome %s\n", buf);
+}
+
+int main(int argc, char *argv[]){
+	func(argv[1]);
+	return 0;
+}

### 2016-01-16 docs/paper.tex
Paper: Finalize draft of background knowledge

@@
 
+\newminted{nasm}{fontsize=\scriptsize,frame=leftline,framesep=2mm,linenos}
+
 \begin{document}
 
-\maketitle
-
+% \maketitle
+%
 % \section*{Abstract}
 % \label{sec:abstract}
-
+%
+% \setcounter{tocdepth}{1}
 % \tableofcontents
-
+%
 % \newpage
@@
 
+\begin{figure}[htpb]
+    \centering
+    \includegraphics[width=\textwidth]{gfx/x86_registers.pdf}
+    \caption{Register overview including \SI{64}{\bit} extension}
+    \label{fig:registers}
+\end{figure}
+
+\Cref{fig:registers} (from
+Wikipedia\footnote{\url{https://en.wikipedia.org/w/index.php?title=X86&oldid=696308590\#/media/File:Table_of_x86_Registers_svg.svg}})
+shows an overview of registers available on the x86 platform. While there are
+dedicated registers for floating pointer operations and also special registers
+which hardware protection (segment registers) we will only focus on nine most
+commonly used registers.
+
+\begin{minipage}[t]{0.5\textwidth}
+    \begin{description}
+        \item[\texttt{EAX}] Accumulator Register
+        \item[\texttt{EBX}] Base Register
+        \item[\texttt{ECX}] Counter Register
+        \item[\texttt{EDX}] Data Register
+        \item[\texttt{ESI}] Source Index
+        \item[\texttt{EDI}] Destination Index
+        \item[\texttt{EBP}] Base Pointer
+        \item[\texttt{ESP}] Stack Pointer
+        \item[\texttt{EIP}] Instruction Pointer
+    \end{description}
+\end{minipage}
+\begin{minipage}[t]{0.45\textwidth}
+    \begin{figure}[H]
+        \includegraphics[width=\textwidth]{gfx/single_register.png}
+        \caption{Addressing specific parts of a register including \SI{64}{\bit} extension}
+        \label{fig:single_register}
+    \end{figure}
+\end{minipage}
+\bigskip
+
+The instruction pointer \texttt{EIP} points to the next instruction located in
+memory which is going to be executed on the cycle. Stack pointer \texttt{ESP} and base
+pointer \texttt{EBP} are used for stack management which is vital to call and
+return from multiple functions properly. The remaining six registers are used
+for computation and passing arguments for system calls. Their values can either
+be interpreted as integers or pointers.
+
+Note that these registers can be addressed partially allowing one to write only
+to the lower \SI{16}{bit} for example as displayed in
+\cref{fig:single_register} taken from \textit{null
+programm}\footnote{\url{http://nullprogram.com/img/x86/register.png} on
+December of 2015}.
+
+The CPU comes with protection mechanisms which allows the operating system
+kernel to limit the privileges of other processes. This mechanism is known as
+\textit{protection rings} (Ring 0 -- Ring 3). The kernel runs \emph{in} Ring 0
+(most privileged) and switches to Ring 3 (least privileged) when a normal
+process is scheduled. A system call has to be made by the process if it needs
+something which goes beyond its scope. The kernel takes over, deals with the
+request and returns execution back to the process. This is known as
+\textit{context switch} and switching between Rings happens along the way.
+
+\subsection{System Calls}
+
+As already mentioned in the previous paragraph, a process only has limited
+capabilities and the kernel has to take over to fulfill certain (more
+privileged) operations. The operating system's documentation tells you which
+system calls are available (on which platform) and what additional parameters
+they require. Let us illustrate this with an example: On x86 the Linux system
+number 4 (starting from 0) is the \texttt{sys\_write} system call which writes
+data to a file descriptor. It takes three arguments, the file descriptor to
+write to, a pointer to the start of the data which should be written and the
+length of the data. The number of the system call together with these three
+parameters are placed in the \texttt{EAX}, \texttt{EBX}, \texttt{ECX},
+\texttt{EDX} respectively. To invoke the system call issue following
+instruction:
+
+\begin{nasmcode*}{autogobble,xleftmargin=1cm,linenos=false}
+    int     0x80
+\end{nasmcode*}
+
+Nowadays you may encounter a different mechanism for system calls, the Virtual
+Dynamic Shared Objects (vDSO) mechanism. This goes beyond our scope here, we
+will use the previously mentioned mechanism in our exploits. If interested, you
+may want to look at the related man
+page\footnote{\url{http://man7.org/linux/man-pages/man7/vdso.7.html}}.
+
 \subsection{Memory}
 
+Physical memory is managed by the operation system kernel by utilising the
+Memory Management Unit (MMU). Each process' address space is virtualized and
+memory operations are translated on-the-fly by the MMU. The physical memory is
+segmented into \textit{pages} (typically \SI{4}{\kibi\byte} in size) and each
+page can be mapped into the virtual address space of one or more (shared)
+processes.~\cite[pp.~400]{unix_interals}
+
+The main parts located inside the (virtual) address space of a process are the
+executable itself with its \texttt{.text} and \texttt{.data} section, the heap
+used for dynamic data, the stack used for local variables and function calling
+and used libraries.
+
+\subsection{Endianness}
+
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=0.35\textwidth]{gfx/little_endian.pdf}
+    \caption{Placement of bytes in memory in little-endian}
+    \label{fig:little_endian}
+\end{figure}
+
+Endianness refers to the byte order used when storing data in memory (or
+transmitting it over the network). \Cref{fig:little_endian} (from
+Wikipedia\footnote{\url{https://en.wikipedia.org/w/index.php?title=Endianness&oldid=696417697\#/media/File:Little-Endian.svg}})
+illustrates that the least significant byte of a word is placed at the lower
+memory address and successive bytes are placed as the memory address increases.
+
+\subsection{Calling Convention}
+
+A calling convention defines how function calls should be implemented. What
+calling convention is used depends on the platform, toolchain and (compiler)
+settings. Let us exhibit what the convention defines and what convention we are
+using.
+
+\begin{minipage}[t]{0.47\textwidth}
+    Defines:
+    \begin{itemize}
+        \item Where to place arguments
+        \item Where to place return value
+        \item Where to place return address
+        \item Who prepares the stack
+        \item Who saves which register
+        \item Who cleans up\\
+            (caller or callee)
+    \end{itemize}
+\end{minipage}
+\begin{minipage}[t]{0.47\textwidth}
+    C Declaration (cdecl):
+    \begin{itemize}
+        \item Arguments on stack (reverse order)\\
+            stack aligned to \SI{16}{\byte} boundary
+        \item Return via register (\texttt{EAX} / \texttt{ST0})
+        \item \texttt{EAX}, \texttt{ECX}, \texttt{EDX} saved by the caller\\
+            rest saved by the callee
+        \item On stack:\\
+            old instruction pointer (\texttt{IP})\\
+            old base pointer (\texttt{BP})
+        \item Caller does the cleanup
+    \end{itemize}
+\end{minipage}
+
 \newpage
+
+\section{Format String Exploits}
+
+\section{Buffer Overflow}
+
+\section{Shell Code}
+
+\section{Data Execution Prevention (DEP)}
+
+\section{Return Oriented Programming (ROP)}
+
+\section{Address Space Layout Randomization (ASLR)}
+
+\section{Stack Cookies}
+
+\section{Control Flow Integrity (CFI)}
+
+\section{Other Architectures}
+
+\section{Conclusion}
 

### 2016-01-16 docs/paper.tex
Paper: Draft format string exploits

@@
 \subsection{Endianness}
+\label{sub:endianness}
 
@@
 \subsection{Calling Convention}
+\label{sub:calling_convention}
 
@@
 
+\section{Format String Exploits}
+
+The first exploitation technique we will discuss builds upon the interpretation
+of format strings. \texttt{printf} is a C function of the standard library
+which will interpret such strings and print them to \texttt{stdout}. As the
+name already tells you, the supplied string contains \textit{formatter}
+describing how to actually handle additional arguments. If you are unfamiliar
+with \texttt{printf} please have a look at the man
+page\footnote{\url{http://linux.die.net/man/3/printf}}.
+
+Taking a closer look at \texttt{printf} we can see that its first argument is a
+format string followed by a variable number of additional arguments. In C you
+don't know how many arguments have been supplied when a function with a
+variable number of arguments is called. Some instances work around this by
+taking an argument count as their first argument, others expect you to
+terminate with a special symbol (usually \texttt{NULL}). \texttt{printf} uses
+the format string to derive how many arguments have been supplied. Calling
+\texttt{printf}, for example, with the string \texttt{"\%d + \%d = \%d"}
+assumes that (at least) three arguments have been provided.
+
+\begin{listing}[h!]
+    \begin{minipage}[t]{0.47\textwidth}
+        \cfile{../format_string/main.c}
+    \end{minipage}
+    \begin{minipage}[t]{0.47\textwidth}
+        \begin{pre*}{autogobble}
+            > echo foobar | ./main
+            You entered:
+            foobar
+        \end{pre*}
+        \medskip
+        \begin{pre*}{autogobble}
+            > echo AAAABBBB | ./main
+            correct
+        \end{pre*}
+        \medskip
+        \begin{pre*} {autogobble}
+            > echo '%08x' | ./main
+            You entered:
+            bfd98ed4
+        \end{pre*}
+    \end{minipage}
+    \caption{Program vulnerable to Format String Exploits}
+    \label{src:format_string_exploit}
+\end{listing}
+
+The exploit comes from the notion that a format string provided by an attacker
+gets interpreted. The program shown in \cref{src:format_string_exploit} will
+take an arbitrary string from \texttt{stdin} and pass it on to \texttt{printf}.
+For simple inputs (not containing formatter) this works fine. But as soon as
+formatter are provided, \texttt{printf} is going to access the locations where
+the corresponding arguments \emph{would} be located. From the calling
+convention described in \cref{sub:calling_convention} we know that these
+arguments would be located on the stack, therefore \texttt{printf} will print
+whatever lies on the stack.
+
+To fully exploit the provided example, note that an attacker in this scenario
+wants to get a hold of the hardcoded password stored in \texttt{passwd}. Since
+local variables are placed on the stack \texttt{printf} will be able to read
+the password if enough formatters are provided:
+
+\begin{pre*} {autogobble}
+    > python -c 'print "%08x." * 10' | ./main
+    bf920c14.00000064.b77de29e.00000000.00000000.b77fedf8.bf920d94.00000000.41414141.42424242.
+\end{pre*}
+
+Here we use Python to craft the format string for us. As we can see the
+password is printed (ASCII encoded). Byte order is swapped because of
+endianness (see \cref{sub:endianness}). Apart from the password we also gather
+a bunch of pointers, these can be used later on to break ASLR (see
+\cref{sub:info_leak}).
+
+We would like to point the reader to the book \textit{Hacking: The Art of
+Exploitation} \cite[pp.~167]{art_of_exploitation} for more details about this
+technique. We will come back to this technique later on to show that
+\texttt{printf} enables even more sophisticated attacks.
+
 \newpage
-
-\section{Format String Exploits}
 

### 2016-01-17 overflow/overflow.c
Paper: Draft buffer overflow exploits

@@ -1,19 +1,17 @@
 #include <stdio.h>
 
-void mordor(){
-	printf("One does not simply jump into mordor()!\n");
+void mordor(void) {
+    printf("One does not simply jump into mordor()!\n");
 }
 
-void echo(){
-	char buffer[20];
-	printf("Enter text:\n");
-	scanf("%s", buffer);
-	printf("You entered: %s\n", buffer);
+void echo(void) {
+    char buffer[20] = {0};
+    printf("Enter text:\n");
+    scanf("%s", buffer);
+    printf("You entered: %s\n", buffer);
 }
 
-int main(){
-
-	echo();
-	
-	return 0;
+int main(void) {
+    echo();
+    return 0;
 }

### 2016-01-18 shellinjection/vuln.c
Reorganize source code and sections

@@ -1,13 +1,0 @@
-#include <stdio.h>
-#include <string.h>
-
-void func(char *name){
-	char buf[100];
-	strcpy(buf, name);
-	printf("Welcome %s\n", buf);
-}
-
-int main(int argc, char *argv[]){
-	func(argv[1]);
-	return 0;
-}

### 2016-01-18 src/buffer_overflow/overflow.c

@@ -1,12 +1,13 @@
 #include <stdio.h>
 
 void mordor(void) {
-    printf("One does not simply jump into mordor()!\n");
+    puts("One does not simply"
+         "jump into mordor()!");
 }
 
 void echo(void) {
     char buffer[20] = {0};
-    printf("Enter text:\n");
+    puts("Enter text:");
     scanf("%s", buffer);
     printf("You entered: %s\n", buffer);
 }

### 2016-01-18 src/shell_code/vuln.c

@@ -1,0 +1,15 @@
+#include <stdio.h>
+#include <string.h>
+
+void func(char *name) {
+    char buf[100] = {0};
+    strcpy(buf, name);
+    printf("Welcome %s\n", buf);
+}
+
+int main(int argc, char *argv[]){
+    if (argc == 2) {
+        func(argv[1]);
+    }
+    return 0;
+}

### 2016-01-22 rop/rop.c
Paper: Draft ROP, ASLR and StackGuard

@@ -1,19 +1,0 @@
-#include <stdio.h>
-#include <string.h>
-
-char* not_used = "/bin/sh";
-
-void not_called(){
-		printf("Not quite a shell...\n");
-		system("/bin/date");
-}
-
-void vulnerable_function(char* string){
-	char buffer[100];
-	strcpy(buffer, string);
-}
-
-int main(int argc, char** argv){
-		vulnerable_function(argv[1]);
-		return 0;
-}

### 2016-01-22 src/return_oriented_programming/rop.c

@@ -1,0 +1,21 @@
+#include <stdio.h>
+#include <string.h>
+
+char *not_used = "/bin/sh";
+
+void not_called(void) {
+    puts("Not quite a shell...");
+    system("/bin/date");
+}
+
+void vulnerable_function(char* string) {
+    char buffer[100] = {0};
+    strcpy(buffer, string);
+}
+
+int main(int argc, char *argv[]) {
+    if (argc == 2) {
+        vulnerable_function(argv[1]);
+    }
+    return 0;
+}

### 2016-01-24 src/return_oriented_programming/rop.c
Paper: Review ROP, ASLR and StackGuard

@@ -1,4 +1,5 @@
 #include <stdio.h>
+#include <stdlib.h>
 #include <string.h>
 
 char *not_used = "/bin/sh";
